Drought-sensitive trees survive on deep water access in a tropical forest community

### 2020-08-03 manuscript.rmd
edited a discussion para

---
title: "Drought-sensitive trees survive on deep water access in a tropical forest community"
author: ''
date: ''
output:
  bookdown::html_document2:
    number_sections: no
    theme: flatly
    fig_caption: yes
  bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
        fig_caption: yes
  bookdown::pdf_document2: default
bibliography: references.bib
csl: new-phytologist.csl
---

@@

\
**Corresponding author**: 
Rutuja Chitra-Tarak
\
E: arutuj@gmail.com
\
T: +1(505) 665 1647

@@

\
Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism. This mechanism assumes that water availability varies belowground by vertical depth and so do rooting depths of co-occuring trees, by species and size. Consequently trees' in the same forest differ in their drought experience, and thus in their responses in term of growth and mortality. A key bottlneck in community-wide testing of this mechanism has been lack of data on trees' rooting or water-sourcing depths, followed by data on water-availability at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the World’s 65 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though ~60-80 % (40 - 53 throusand species) of the World’s tree diversity resides in the tropics [@Slik:2015]. Use of stable isotopes of water as a tracer provides an indirect way of trees' water sourcing depths by matching isotopic value in plant tissue to variations in soil pore water vertically belowground. Such data are available only for ~0.05% of World's tree species [@Evaristo:2016]. In general, community-scale data collection for rooting or water-sourcing depths is especially challenging for species-rich tropical forests. Further, to estimate plant available water with soil depth requires hydrological data and modeling, but these are rarely co-located with forest inventory sites (but see @Johnson:2018fad;@Kumagai:2011;@Ivanov:2012)

\
Establishing relationships between plant's environment, traits and demographic outcomes is imperative for a predictive plant ecology [@Mcgill.2006], but plant's realised water environments are rarely studied. Huge efforts are devoted to linking the same-for-all-trees drought indices, above-ground traits and demographic rates. These efforts have resulted in identification of above-ground traits such as proxies of hydraulic failure as indicators of mortality risk. A recent meta-analysis found greater drought-induced mortality rates in species with cavitation-vulnerable xylem ($\Psi_{88,~stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~stem}$) [@Anderegg:2016], but counter examples also exist [@Hoffmann:2011; @Paddock:2013]; indicating whether plants reach critical thresholds of hydraulic failure depends on their realised drought intensity (see Table \@ref(tab:symbols) for trait symbols and their definitions). A rare study that accounted for realised water availability for four tree species in a centennial drought (Texas, USA) found that mortality decreased with deep rootedness even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi_{88,~stem}$) and had narrower safety-margins ($\Psi_{min} - \Psi_{88,~stem}$) [@Johnson:2018fad]. A species sourcing deep water in a Karst system (Italy) showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi_{88,~stem}$ and narrower safety margin compared to two species sourcing shallow-water [@Nardini:2015]. @Chitra-Tarak:2018 proposed the first community-scale model that linked trees' realised water availability to demographic outcomes through droughts. They found that species accessing deep-water in a tropical dry forest (India) died more in a rare, centennial drought compared to shallow-water species. They concluded that this was consistent with an 'aquisitive-conservative' species strategy spectrum. Aquisitives relied on stable water resource at depth, which exhausted in the multi-annual drought, resulting in reduced survival, whereas conservative species used fluctuating shallow-water and survived the drought as some shallow-water was available via infiltration. However, lack of data on rooting/water-sourcing depths and hydraulic traits precluded this study from corroborating modelled estimates of water-uptake depths, or investigating trait coordination.

\
Trade-offs that shape how trees' make a living in their environment (ecological strategies) provide Dynamic Vegetation Models (DVMs) the essential constraints to represent species-rich communities under changing environment [@Koven:2020]. This entails indentifying trade-offs among environment, traits and demography. A universal 'fast-slow' (aka productive-persistent, or acquisitive-conservative) plant economics spectrum conceptualises plants coordination for all resources and organs, both above- and below-ground [@Reich:2014]. For water resource observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified a fast-slow spectrum, although examples of all trade-offs observed simultaneously are rare, and belowground trait or environment linkages are weak. 'Fast' species at the productive end of the spectrum have traits with high capacity to acquire, move, store and use water (high $K_{max}$, low wood density) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resources, whereas species with slow traits are better protected from C losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{tlp}$, larger safety margin). Fast strategy is associated with spatially wet sites such as valleys or high rainfall [@Kursar:2009; @Oliveira:2019], but whether and how the fast-slow spectrum holds true for the moisture gradient in the soil column is not clear. While deep rootedness may be expected to support a fast strategy with a steady water supply unlike shallow-rooted trees. However, this may depend on the exact dynamics of water-availability by depth at a site, depending on precipitation regime and hydrology. In pulse-driven arid systems *shallow*-rooted plants are associated with fast traits (high $K_{max}$) that maximise the use of the water-pulse, while *deep*-rooted plants have slow traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013].

\
Here, for large trees of xx canopy species of a hyperdiverse neo-tropical moist forest (BCI, Panama) we make a thorough assessment of trade-offs between belowground water environment, above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 35 years that withnessed droughts of various intensity and duration. We obtained daily dynamics of plant water potential ($\Psi_{soil,~z}$) by depth $z$ in the whole-soil column with a locally parametarised and calibrated 1-D hydrological water balance model. Developing a novel inverse model allowed us to estimate species effective rooting depths (ERD), defined as the depth ($z$) whose soil water dynamics ($\Psi_{~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015). This model incorporated impact of atmospheric drought (VPD) on growth and was constrained by above-ground hydraulic traits, namely, species leaf vulnerability curves ($K_{leaf}$ vs. $\Psi_{leaf}$). We collected new data on vulnerability curves for xx species. Identifying scaling relationships between these curves and commonly available traits (SLA, specific wood density) allowed us to scale-up these curves to `r length(data.model.AB$sp)` tree species in the community. We evaluated ERD against existing data on plant and soil water isotopes ($\delta^2H$) and coordination of ERD with hydraulic traits and mortality rates.

@@

## Study Site Description
\
We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterisation in terms of long-term forest demography, plant functional ecology, hydrology, and water sourcing depth water isotopes on tree species and soil from an independent isotopic tracer study on trees’ water-sourcing depths [@Meinzer:1999]. The 1,560 hectare island was formed after the river Chagres was dammed in 1914 that created Gatun Lake, the main channel of the Panama Canal. Rainfall at BCI is seasonal with a mean annual rainfall of 2687 mm (± 515 SD over 1983-2017) and a four-month long dry season from mid-December through mid-April (≤ 100 mm rain/month). The entire island is forested and classified as a tropical moist forest. Although percentage of canopy species that are at least partly deciduous during the dry season is 28%, the maximum percentage of deciduous individuals in the canopy is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% [@Condit:2000]. Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014].

\
In the 50 ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species level and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm size class [@Condit:2017]. This inventory represents 321 woody species across the eight censuses. Aside from a 2 ha patch of ~100 year old secondary forest at the north-eastern corner of the plot the remaining 48 ha represent old growth forest that was never clear cut for agriculture. The plot (100 m x 50 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001]. The plot covers most of the upper catchment of Conrad Creek. The shallow valley of this creek is the main interruption to the otherwise mostly flat or gently sloping plot. The valley gives rise to the 1.5-ha swamp at the end of the wet season. Soils of the plot are homogenous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot (BCI soil survey report). The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil. Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding. Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere.

@@

where, $q_{rain}$ is rainfall, $E_{v}$ is evapotranspiration (ET) of the forest, $E_{g}$ is ground evaporation, $q_{over}$ is surface runoff, $q_{h2osfc}$ is runoff from surface water storage, $q_{drai}$ is sub-surface drainage (all in mm H~2~O) and $\Delta t$ is the time-step. $z$ represents the 15 soil layers exponentially distributed in depth, here, (r soil.depths) (m).

@@

We ran ELM-FATES for the 5000 parameter ensembles with hourly climate drivers from 1985-2018 initialised with the stand structure of the 50 ha plot in a single site mode ($0.1^\circ$ x $0.1^\circ$ centered at $9.1530^\circ$N, $280.2539^\circ$E). As our key interest here was in soil water availability rather than forest dynamics, we ran ELM-FATES in simplified modes: static stand structure and without plant hydrodynamics, and with a single Plant Functional Type (PFT) of evergreen trees. The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of dry-deciduous PFT did not significantly alter results (not shown).

@@

```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE; m) versus dry season stable isotopic concentration $\\delta^2H$ (mean $\\pm$ (add error); ‰) for ", length(unique(erd.iso$sp))," canopy species from Barro Colorado Island. Data from [@Meinzer:1999]. See Table  \\@ref(tab:spcodes) for scientific names corresponding to species codes.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_phenology_Meinzer_gr.Psi.VPD.jpeg"))
```

@@

We calculated mortality rates for large trees (>= 10 cm dbh) of species with ERD estimates (all canopy species with maximum height >= 30 m) and average adbunance of at least 20 trees (one species was eleminated with an average abundance of 9 trees; *n* = `r length(erd.mrate.sp)`). For each of the seven census intervals ($t$) in the 33 year record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successives censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{t}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50 ha plot of a species present in cen~1~ and dead in cen~2~, respectively, and t is the duration based on mean dates of cen~1~ and cen~2~. We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{i=1}^{6} {Mort_{t}}$ and, finally, species mortality anoamalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$. To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$). To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~leaf}$.

@@

\
Species with deeper $ERD$ showed lower mean mortality rates across the 33 year record (1982-2015; *R*^2^ = 0.13, *P* = 0.029, *n* = `r length(erd.mrate.sp)`; Figure \@ref(fig:erd-mort)). For five out of seven census intervals, species with shallower $ERD$ had significantly higher mortality anomalies than deeper ERD species; Figure (\@ref(fig:erd-mort-anom)). Analysis of time spent under catastrophic soil drought showed that the proportion of days species $\Psi_{ERD}$ was more negative than species $\Psi_{crit}$ decreased with species ERD (Figure \@ref(fig:erd-time)). Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25 year period for which soil water potential $\Psi_{soil,~z}$ was modelled (1990-2015), as soil depths at and below 1.7 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dryseason and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).

@@

# Discussion
\
Our analyses revealed that trees with deeper effective rooting depths (ERD) were associated higher hydraulic efficiency, but not safety: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species. Yet, deeper ERD species showed lower mortality rates than shallower ERD species over the 33 year record (1982-2015). Analysis of time spent below ($\Psi_{crit}$) in the soil zone matching ERD $\Psi_{soil,~z = ERD}$ showed that this was likely due to decreasing realised risk of hydraulic failure with depth. We thus show, for the first time, that interspecific variation in trees' effective rooting depths aligns species on a 'fast-slow' water economic spectrum by demonstrating a link between trees' realised drought experience, hydraulic architecture and long-term mortality outcomes. On the 'fast' end of the spectrum were deep ERD species with hydraulically efficient, but risky suit of traits optimised for an environment of reliable and abundant source of water at depth that actualised to a lower mortality-risk. At the slow end of the spectrum were shallow ERD species with inefficient, but safer xylem, likely optimised for high variability in $\Psi_{soil}$ in the shallow soil layers, resulting in higher variance in mortality rates across censuses as $\Psi_{soil,~z = ERD}$ crossed ERD $\Psi_{crit}$. Mortality rates of shallow ERD species may have been exacerbated by increased drought events during the study period. Previous work shows that half of the extreme dry seasons, estimated as < 1 SD annual water deficit, during 1931-2015 (84 years) occurred during the first 23 years (1982-2005) of the study period. [@Condit.2017]

### Effective rooting depths at tree community level
\
By estimating ERD for large trees of 33 canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species species-rich tropical forests. We note the availability of data on $\delta^2^H$ on trees and water sources belowground--soil or substrate matrix water in the unsaturated vadoze zone or free water below the water table--as important for model calibration/validation. The first inverse model of ERD (water uptake depth index, *n* = 12 species) developed by Chitra-Tarak et al @Chitra-Tarak:2018 lacked validation as existing methods were inadequate to extract water bound to the soil and saprolite found in their dry forest study site (Mudumalai National Park, Southern India; Jean Riotte, pers. comm.). Our ERD model was distinctly constrained on species leaf vulnerability (PLC) curves. We benefitted from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (SLA~lamina~ and Wood Specific Gravity) and parameters of PLC curves for which we collected data on a few species (*n* = 14), a 'hard' trait. This allowed us to scale-up PLC curves at the community scale (*n* = 284 species). Whether such scaling relationships between hard and soft traits hold true for other forests and their wider applicability remains to be seen as data on PLC curves in species-rich, tropical forests are limited, but counter examples exist [see xx, yy, zz (Lawren, could you suggest references?].

>>>>

---
title: "Drought-sensitive trees survive on deep water access in a tropical forest community"
author: ''
date: ''
output:
  bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
        fig_caption: yes
  bookdown::html_document2:
    number_sections: no
    theme: flatly
    fig_caption: yes
  bookdown::pdf_document2: default
bibliography: references.bib
csl: new-phytologist.csl
---

@@

\
**Corresponding author**: 
Rutuja Chitra-Tarak, A mispelt word.
\
E: arutuj@gmail.com
\
T: +1(505) 665 1647

@@

\
Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism. This mechanism assumes that water availability varies belowground by vertical depth and so do rooting depths of co-occuring trees, by species and size. Consequently trees' in the same forest differ in their drought experience, and thus in their responses in term of growth and mortality. A key bottlneck in community-wide testing of this mechanism has been lack of data on trees' rooting or water-sourcing depths, followed by data on water-availability at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the World’s 65 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though ~60-80 % (40 - 53 throusand species) of the World’s tree diversity resides in the tropics [@Slik:2015]. Use of stable isotopes of water as a tracer provides an indirect way of trees' water sourcing depths by matching isotopic value in plant tissue to variations in soil pore water vertically belowground. Such data are available only for ~0.05% of World's tree species [@Evaristo:2016]. In general, community-scale data collection for rooting or water-sourcing depths is especially challenging for species-rich tropical forests. Further, to estimate plant available water with soil depth requires hydrological data and modeling, but these are rarely co-located with forest inventory sites (but see, @Johnson:2018fad; @Kumagai:2011; @Ivanov:2012)

\
Establishing relationships between plant's environment, traits and demographic outcomes is imperative for a predictive plant ecology [@Mcgill.2006], but plant's realised water environments are rarely studied. Huge efforts are devoted to linking the same-for-all-trees drought indices, above-ground traits and demographic rates. These efforts have resulted in identification of above-ground traits such as proxies of hydraulic failure as indicators of mortality risk. A recent meta-analysis found greater drought-induced mortality rates in species with cavitation-vulnerable xylem ($\Psi_{88,~stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~stem}$) [@Anderegg:2016], but counter examples also exist [@Hoffmann:2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013]; indicating whether plants reach critical thresholds of hydraulic failure depends on their realised drought intensity (see Table \@ref(tab:symbols) for trait symbols and their definitions). A rare study that accounted for realised water availability for four tree species in a centennial drought (Texas, USA) found that mortality decreased with deep rootedness even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi_{88,~stem}$) and had narrower safety-margins ($\Psi_{min} - \Psi_{88,~stem}$) [@Johnson:2018fad]. A species sourcing deep water in a Karst system (Italy) showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi_{88,~stem}$ and narrower safety margin compared to two species sourcing shallow-water [@Nardini:2015]. @Chitra-Tarak:2018 proposed the first community-scale model that linked trees' realised water availability to demographic outcomes through droughts. They found that species accessing deep-water in a tropical dry forest (India) died more in a rare, centennial drought compared to shallow-water species. They concluded that this was consistent with an 'aquisitive-conservative' species strategy spectrum. Aquisitives relied on stable water resource at depth, which exhausted in the multi-annual drought, resulting in reduced survival, whereas conservative species used fluctuating shallow-water and survived the drought as some shallow-water was available via infiltration. However, lack of data on rooting or water-sourcing depths and hydraulic traits precluded this study from corroborating modelled estimates of water-uptake depths, or investigating trait coordination.

\
Trade-offs shape how trees' make a living in their environment (ecological strategies) and also provide Dynamic Vegetation Models (DVMs) the essential constraints to represent species-rich communities under changing environment [@Koven:2020]. This entails indentifying trade-offs among environment, traits and demography. A universal 'fast-slow' (aka productive-persistent, or acquisitive-conservative) plant economics spectrum conceptualises plants coordination for all resources and organs, both above- and below-ground [@Reich:2014]. For water resource observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified a fast-slow spectrum, although examples of all trade-offs observed simultaneously are rare, and belowground trait or environment linkages are weak. 'Fast' species at the productive end of the spectrum have traits with high capacity to acquire, move, store and use water (high $K_{max}$, low wood density) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resources, whereas species with slow traits are better protected from C losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{tlp}$, larger safety margin). Fast strategy is associated with spatially wet sites such as valleys or high rainfall [@Kursar:2009; @Oliveira:2019], but whether and how the fast-slow spectrum holds true for the moisture dynamics vertically belowground is not clear. Deep rootedness may be expected to support a fast strategy with a steady water supply unlike shallow-rooted trees. However, this may depend on the exact dynamics of water-availability by depth at a site, depending on precipitation regime and hydrology. In contrast, in pulse-driven arid systems *shallow*-rooted plants are associated with fast traits (high $K_{max}$) that maximise the use of the water-pulse, while *deep*-rooted plants show slow traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013].

\
Here, for large trees of length(erd.sp) canopy species of a hyperdiverse neo-tropical moist forest (BCI, Panama) we make a thorough assessment of trade-offs between belowground water environment, above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 34 years that withnessed droughts of various intensity and duration. We obtained daily dynamics of plant water potential ($\Psi_{soil,~z}$) by depth $z$ in the whole-soil column with a locally parametarised and calibrated 1-D hydrological water balance model. Developing a novel inverse model allowed us to estimate species effective rooting depths (ERD), defined as the depth ($z$) whose soil water dynamics ($\Psi_{~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015). This model incorporated impact of atmospheric drought (VPD) on growth and was constrained by above-ground hydraulic traits, namely, species leaf vulnerability curves ($K_{leaf}$ vs. $\Psi_{leaf}$). We collected new data on vulnerability curves for xx species. Identifying scaling relationships between these curves and commonly available traits (SLA, specific wood density) allowed us to scale-up these curves to `r length(data.model.AB$sp)` tree species in the community. We evaluated ERD against existing data on plant and soil water isotopes ($\delta^2H$) and coordination of ERD with hydraulic traits and mortality rates.

@@

## Study Site Description
\
We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and water sourcing depth water isotopes on tree species and soil from an independent isotopic tracer study on trees’ water-sourcing depths [@Meinzer:1999]. The 1,560ha island was formed after the river Chagres was dammed in 1914 that created Gatun Lake, the main channel of the Panama Canal. Rainfall at BCI is seasonal with a mean annual rainfall of 2687 mm (± 515 SD over 1983-2017) and a four-month long dry season from mid-December through mid-April (≤ 100 mm rain/month). The entire island is forested and classified as a tropical moist forest. Although percentage of canopy species that are at least partly deciduous during the dry season is 28%, the maximum percentage of deciduous individuals in the canopy is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% [@Condit:2000]. Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014].

\
In the 50ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species level and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm size class [@Condit:2017]. This inventory represents 321 woody species across the eight censuses. Aside from a 2ha patch of ~100 year old secondary forest at the north-eastern corner of the plot the remaining 48 ha represent old growth forest that was never clear cut for agriculture. The plot (100 m x 50 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001]. The plot covers most of the upper catchment of Conrad Creek. The shallow valley of this creek is the main interruption to the otherwise mostly flat or gently sloping plot. The valley gives rise to the 1.5-ha swamp at the end of the wet season. Soils of the plot are homogenous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot (BCI soil survey report). The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil. Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding. Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere.

@@

where, $q_{rain}$ is rainfall, $E_{v}$ is evapotranspiration (ET) of the forest, $E_{g}$ is ground evaporation, $q_{over}$ is surface runoff, $q_{h2osfc}$ is runoff from surface water storage, $q_{drai}$ is sub-surface drainage (all in mm H~2~O) and $\Delta t$ is the time-step. $z$ represents the depths of exponentially distributed 15 soil layers, here, `r paste(soil.depths, collapse=", ")` (m).

@@

We ran ELM-FATES for the 5000 parameter ensembles with hourly climate drivers from 1985-2018 initialised with the stand structure of the 50ha plot in a single site mode ($0.1^\circ$ x $0.1^\circ$ centered at $9.1530^\circ$N, $280.2539^\circ$E). As our key interest here was in soil water availability rather than forest dynamics, we ran ELM-FATES in simplified modes: static stand structure and without plant hydrodynamics, and with a single Plant Functional Type (PFT) of evergreen trees. The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of dry-deciduous PFT did not significantly alter results (not shown).

@@

```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ add error; m) versus dry season stable isotopic concentration $\\delta^2H$ (mean $\\pm$ 1SE; ‰) for ", length(unique(erd.iso$sp))," canopy species from Barro Colorado Island. Data from [@Meinzer:1999]. See Table  \\@ref(tab:spcodes) for scientific names corresponding to species codes.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_phenology_Meinzer_gr.Psi.VPD.jpeg"))
```

@@

We calculated mortality rates for large trees (>= 10 cm dbh) of species with ERD estimates (all canopy species with maximum height >= 30 m) and average adbunance of at least 20 trees (one species with an average abundance of 9 trees was eliminated; *n* = `r length(erd.mrate.sp)`). For each of the seven census intervals ($t$) in the 34 year record (1982, 1985, 1990, 2000, 2005, 2010, 2015, 2015), species $Mort_{t}$ (%yr^-1^) for two successives censuses (cen~1~ and cen~2~) was calculated as $Mort_{t} = \frac{D_{cen_{2}}}{N_{cen_{1}}}.\frac{100}{t}$, where $N_{cen_{1}}$ and $D_{cen_{1}}$ are the total number of trees in the 50ha plot of a species present in cen~1~ and dead in cen~2~, respectively, and t is the duration based on mean dates of cen~1~ and cen~2~. We calculated species mean mortality rate across all intervals as $Mort_{mean} = \frac{1}{6}\sum_{i=1}^{6} {Mort_{t}}$ and, finally, species mortality anoamalies by intervals as $Mort_{anomaly}$ = $Mort_{t}$ - $Mort_{mean}$. To test whether $ERD$ plays a role in alleviating mortality risk, we regressed species $ERD$ against species ($Mort_{mean}$). To assess whether this may be due to deeper $ERD$ species escaping the experience of catastrophic soil drought, we calculated the proportion of days over the study period (1990-2015) for which soil water potential in the soil zone matching species ERD ($\Psi_{ERD}$) were below species $\Psi_{crit}$, defined as $\Psi_{80,~leaf}$.

@@

\
Species with deeper $ERD$ showed lower mean mortality rates across the 34 year record (1982-2015; *R*^2^ = 0.11, *P* = 0.075, *n* = `r length(erd.mrate.sp)`; Figure \@ref(fig:erd-mort)). For five out of seven census intervals, species with shallower $ERD$ had significantly higher mortality anomalies than deeper ERD species; Figure (\@ref(fig:erd-mort-anom)). Analysis of time spent under catastrophic soil drought showed that the proportion of days species $\Psi_{ERD}$ was more negative than species $\Psi_{crit}$ decreased with species ERD (Figure \@ref(fig:erd-time)). Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25 year period for which soil water potential $\Psi_{soil,~z}$ was modelled (1990-2015), as soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dryseasons and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).

@@

# Discussion
\
Our analyses revealed that trees with deeper effective rooting depths (ERD) were associated higher hydraulic efficiency, but not safety: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species. Yet, deeper ERD species showed lower mortality rates than shallower ERD species over the 34 year record (1982-2015). Analysis of time spent below ($\Psi_{crit}$) in the soil zone matching ERD $\Psi_{soil,~z = ERD}$ showed that this was likely due to decreasing realised risk of hydraulic failure with depth. We thus show, for the first time, that interspecific variation in trees' effective rooting depths aligns species on a 'fast-slow' water economic spectrum by demonstrating a link between trees' realised drought experience, hydraulic architecture and long-term mortality outcomes. On the 'fast' end of the spectrum were deep ERD species with hydraulically efficient, but risky suit of traits optimised for an environment of reliable and abundant source of water at depth that actualised to a lower mortality-risk. At the slow end of the spectrum were shallow ERD species with inefficient, but safer xylem, likely optimised for high variability in $\Psi_{soil}$ in the shallow soil layers, resulting in higher variance in mortality rates across censuses as $\Psi_{soil,~z = ERD}$ crossed ERD $\Psi_{crit}$. Mortality rates of shallow ERD species may have been exacerbated by increased drought events during the study period. Previous work shows that half of the extreme dry seasons, estimated as < 1 SD annual water deficit, during 1931-2015 (85 yrs) occurred during 1982-2005 (24 yrs) of the study period (1982-2015). [@Condit.2017]

### Realised drought risk integral to assessing drought-vulnerability

Our study brings attention to the need for assessing drought-sensitivity in terms of species realised drought risk. and the usefuless of looking at species ecological strategies for that. As those are developed in response to plants historical environemnt. Plant traits hold clue for the historical environment that shaped their ecological strategy [@Reich:2014].

Our result of lower mortality in deep rooted trees, through a study period of greater recurrence of drought, is consistent with other studies, where among the three or four species studied through a centennial (Texas) or extreme (Italy) drought, deep rooted species showed lower mortality than shallow rooted species. Our result contrasts with the finding of @Chitra-Tarak:2018 who found that species identified as effectively deep rooted by their inverse model died *more* in a prolonged, centennial drought. Their hydrological model found that in the multi-annual drought deep water was exhausted for over two years, whereas our hydrological modeling at BCI finds that despite higher frequncy of droughts during our study period $\Psi_{soil}$ below the depth of 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dryseasons and drought periods, that is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa). The difference in our results are likely attributable to site differences in the precipitation regime. In their dry forest site (Mudumalai National Park, India) precipitation is low (1095 $\pm$ 296 mm; mean $\pm$ SD) compared to BCI (2687 $\pm$ 515 mm), and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (PET = 9xx $\pm$ xx mm), and thus at BCI deep soil layers are effectively recharged annually [@Chitra-Tarak:2018 and this study]. We note that the variation in mortality explained by in our study is low (*R*^2^ = 0.11, *P* = 0.075), and other causes may explain observed mortality. For example. based on a year-long study, lightening explained ~4.5% of mortality for trees >=10 cm dbh and ~40% for trees >=60 cm dbh [@Yanoviak:2019].

### Effective rooting depths at the tree community level
\
By estimating ERD for large trees of length(erd.sp) canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species species-rich tropical forests. Our chosen ERD model is an advancement over the only other inverse model of ERD (or water uptake depth index, @Chitra-Tarak:2018, *n* = 12 species) in terms of physiologically meaningful representation, fewer parameters, corroboration with tree and soil data on $\delta^2H$. Model by @Chitra-Tarak:2018 estimated six species-specific phenomenological parameters: three as root fraction for corresponding lumped vadose zone layers, and three as shape parameters for water uptake vs availability curves. For each species, they calculated daily water uptake successively from three layers--the supply, to satisfy the demand of the daily maximum evapo-transpiration, a function of PET and LAI. They regressed total supply to total demand ratio over census intervals (termed as growth coefficient) against growth (5 data points), and chose values for six parameters that best-fit (highest R^2^, that is, 2 additional parameters). Our inverse ERD model fits only two parameteres by way of regressing depth-specific model against 5 growth data points. Our model was constrained on VPD and species leaf vulnerability (PLC) curves, variables that are shown to be physiologically meaningful [add references]. We benefitted from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (SLA~lamina~ and Wood Specific Gravity) and parameters of PLC curves for which we collected data on a few species (*n* = 14), a 'hard' trait. This allowed us to scale-up PLC curves at the community scale (*n* = 284 species). Whether such scaling relationships between hard and soft traits hold true for other forests and their wider applicability remains to be seen as data on PLC curves in tropical forests are limited, but counter examples exist [see xx, yy, zz (Lawren, could you suggest references?]. We also note the availability of data on $\delta^2^H$ on trees and water sources belowground--soil/substrate matrix water in the vadoze zone or free water below the water table--as important for model calibration/validation. Model by Chitra-Tarak et al @Chitra-Tarak:2018 were not corroborated with such data as existing methods were inadequate to extract water bound to the soil and saprolite found in their dry forest study site (Jean Riotte, pers. comm.). Our model of belowground water-availability (ELM-FATES) was also advanced compared to that used by Chitra-Tarak et al [2018] and was calibrated with data on soil moisture dynamics by depth (15, 40, 100 cm), as well as, evapotranspiration and stream discharge, unlike theirs, which was calibrated on stream discharge and water table levels. We lacked data on water table levels which would have been useful to constrain water availability dynamics in the unobserved soil layers.
